Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-10 doc/mc-paper.tex
Incorporated MICE data; added graphs to paper

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{d4}}
\newcommand{\Dfour}{\progname{d4}}
\newcommand{\generator}{\progname{d2p}}
\newcommand{\Generator}{\progname{d2p}}
\newcommand{\checker}{\progname{crat-check}}
\newcommand{\Checker}{\progname{Crat-check}}

@@

\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

\newcommand{\program}[1]{\textsc{#1}}
\newcommand{\lean}{\program{lean4}}

@@

\begin{abstract}
\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model
count. Existing knowledge compilers, however,
provide no guarantee that the representation they generate
is logically equivalent to the original formula, and therefore they cannot validate the generated counts.

@@

We have developed a program that can generate POG representations
from the d-DNNF graphs generated by the state-of-the-art knowledge
compiler D4, as well as independently checkable proofs that the
output POGs are equivalent to the input CNF formulas.
We have implemented an efficient checker
that can handle very large proofs. Our tool
chain for generating and verifying POGs scales to all but the largest d-DNNF graphs
generated by D4 for formulas
from a recent model counting competition. Moreover, we
present a checker that has been formally verified
in the Lean~4 proof assistant, thereby creating the
first formally verified toolchain for weighted and unweighted model
counting.

%  Knowledge compilers convert Boolean formulas into a form 
%  that enables efficient evaluation of properties of the formula,
%  including
%  weighted and unweighted model counting.
% Existing knowledge compilers, however,
%provide no guarantee that the generated representation accurately
%describes the original formula.
%
%We present the Partitioned Operation Graph (POG) representation that
%generalizes existing knowledge compiler representations, and a proof
%framework that can prove the equivalence of a POG representation with
%a conjunctive normal form (CNF) Boolean formula.
%
%We have developed a program that can generate POG representations and
%equivalence proofs from the d-DNNF graphs generated by a leading
%knowledge compiler.  We also present two proof
%checkers---a fast checker that can handle the large proofs generated,
%and a formally verified checker written in the \lean{} proof system.
%Our generation and checking tool chain scales to all but the largest formulas from a
%recent model counting competition.  
\end{abstract}

@@

@@

\begin{definition}[Ring Evaluation Problem (REP)]
  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $1-w(x)$.

@@

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow 
indepdently assigning weights $W(x)$ and $W(\obar{x})$ to each variable $x$, potentially having
$W(x) + W(\obar{x}) \not = 1$.
We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi, w)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{description}
\item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
\item[Partitioned Product] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
\item[Partitioned Sum] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
\end{description}
\end{proposition}

@@

\caption{Example formula (A), its POG representation (B), and its CRAT proof (C)--(F)}

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a terminal node with literal $x$ or $\nodeu_1$ is a
  product node containing $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a terminal node with literal $\obar{x}$ or $\nodeu_0$ is a product node containing $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined, via two proof steps, with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal 
  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then importing the generated unsatisfiability prove
  as assertions in the CRAT file.
\item If there is a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$, 
\item If there are multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
\end{enumerate}
  We then combine as hints the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A))
to generate the target clause for node $\nodeu$.
\end{enumerate}
Observe that most of these steps involve a polynomial number of
operations per recursive call.  The only potential source of
exponential growth is when a SAT solver must be invoked to validate a
literal.

@@

Completing the equivalence proof of input formula $\phi$ and its POG
representation with root literal $r$ requires showing that
$\modelset(P) \subseteq \modelset(\phi)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$ and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\psi$.  Deleting $C$ requires proving that
$\modelset(\psi) = \modelset(\psi \lor C)$, i.e., that deleting $C$
does not make the formula easier to satisfy.

@@

Each input clause deletion can be done as a single RUP step.  The
proof generator builds a sequence of hints from among the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  A RUP proof for input clause $C$ effectively proves that any
total assignment $\assign$ that does not satisfy $C$ will yield
$\assign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} | \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

Working upward through the graph, each node $\nodeu$ is {\em marked} when
the collected set of literals force $u$ to evaluate to $0$.  The
program adds $\obar{u}$ to the RUP literals and adds the appropriate
defining clause to the list of hints.  A terminal node with literal
$\lit$ will be marked if $\lit \in C$, with no hint required.  If
product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
added to the hints.  Marking sum node $\nodeu$ requires that both of
is children $\nodeu_1$ and $\nodeu_2$ be marked.  The first defining
clause for this node (see Table~\ref{tab:defining}(B)) will then be added.  At the very end, the program
(assuming the reverse implication holds) will attempt to mark root
node $\noder$, which would require $\assign(r) = 0$, yielding a
conflict.

It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
each clause deletion requires a single RUP step, where 
the number of hints will be less than the number of
nodes in the POG\@.

@@

The performance of forward implication, both in its execution time and
the size of the proof generated by two optimizations described here.
A key feature of both is that they do not require any changes to the
proof framework---they build on the power of extended resolution to
create new logical structures within the proof.

@@

A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literal as children.  As described
earlier, the standard approach for each child literal $\lit \in \L$ is to first attempt BCP, but calling a SAT solver
on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$.

An alternative approach is to collect those literals $\lit_1, \lit_2,
\ldots, \lit_m$ that cannot be validated by BCP and then define a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that this product must yield 1 for extension
variable $v$.  This can be done by contradiction using a SAT solver.
In particular, looking at the defining clauses for $\nodev$ (Table
\ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
  the first defining clause to simplify to $C_p = \obar{lit}_1 \lor
  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
  clauses to become tautologies.  The formula
  $\simplify{\psi}{\passign} \cup \{ C_p \}$ should therefore be
  unsatisfiable.  The proof steps and hints generated by the SAT
  solver can mapped back to the input clauses, plus the defining
  clauses for $\nodev$.

@@

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.  We can exploit the property
that a shared subgraph in the graph generated by a top-down d-DNNF
compiler arises when the program detects that the simplified set of
clauses arising from a partial assignment to the literals matches that
of a previous traversal.  When this d-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and later check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The forward implication proof can exploit this property of shared
subgraphs by constructing and proving a {\em lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma$ must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.  The net effect is to allow $\validate$ to
maintain the graph structure of the POG, invoking $\validate$ on each
node at most once.

The challenge for implementing this strategy is to find a way to
represent the clauses in a simplified formula $\gamma$.  Some may be
unalterered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CRAT file to create
the appropriate clauses.  Imagine that there is an {\em argument} clause
$C \in \gamma$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  If we
define product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
Table~\ref{tab:defining}(A), we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

Suppose for every clause $C_j \in \gamma$ that does not correspond to
an input clause, we generate a synthetic clause $C'_j$ with guard literal
$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
and
let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
Invoking $\validate(\nodeu, \lassign, \gamma)$
 will then prove a lemma, given by the target clause
 $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
 showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u)$.

>>>>

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\drational}{\mathbb{Q}_{2,5}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\generator}{\progname{d2p}}
\newcommand{\Generator}{\progname{D2p}}
\newcommand{\checker}{\progname{crat-check}}
\newcommand{\Checker}{\progname{Crat-check}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}

\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}

@@

\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

\newcommand{\program}[1]{\textsc{#1}}
\newcommand{\lean}{Lean~4}

@@

\begin{abstract}

\emph{Knowledge compilation} is the process of converting a Boolean 
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model 
count. Existing knowledge compilers, however,
provide no guarantee that the representation they generate
is logically equivalent to the original formula, and therefore they cannot validate the generated counts.

@@

We have developed a program that can generate POG representations 
from the d-DNNF graphs generated by the state-of-the-art knowledge 
compiler \dfour, as well as independently checkable proofs that the 
output POGs are equivalent to the input CNF formulas.
We have implemented an efficient checker
that can handle very large proofs. Our tool
chain for generating and verifying POGs scales to all but the largest d-DNNF graphs
generated by \dfour for formulas
from a recent model counting competition. Moreover, we 
present a checker that has been formally verified
in the \lean{} proof assistant, thereby creating the
first formally verified toolchain for
% weighted and unweighted
model counting.
\end{abstract}

@@

Modern Boolean satisfiability (SAT) solvers provide the ability to
find a satisfying assignment to a Boolean formula or to prove that no
such assignment exists.  The have applications across a variety of
domains, including formal verification of hardware, software, and
security protocols, computational mathematics, and combinatorial
optimization.  Some problems of interest, however, require going
beyond standard Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments, often in cases where there are far to many satisfying
assignments to enumerate individually.  Model counting, in turn, has
applications in artificial intelligence, computer security, and
statistical sampling.  A variant on standard model counting is {\em
  weighted model counting}, where a weighting function is defined for
each assignment and the goal becomes to compute the sum of the weights
of the satisfying assignments.

Model counting is a challenging problem---more challenging that the
already NP-hard problem of Boolean (un)satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal becomes model counting and not just
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters have developed that scale to very large formulas, as
witnessed by their progress in the annual model counting competitions.

One approach to model counting, known as {\em knowledge compilation}
is to transform the formula into a structured form for which model
counting becomes more tractable.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a Boolean formula as a
directed acyclic graph, with terminal nodes labeled by Boolean
variables and their complements, and with each nonterminal node
labeled by Boolean operation And or Or.  Restrictions are place on the
structure of the graph such that a count of the models can be computed
by a single bottom-up traversal of the graph.

Once a Boolean formula has been transformed into a d-DNNF
representation, a number of useful properties of the formula can
readily be computed.  Kimmig, et al., present a very general
framework, known as {\em algebraic model
counting}~\cite{kimmig:jal:2017}, that go well beyond simple model counting.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are corrected, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers are able to produce
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has
shown that adding this capability has also greatly improved the quality of
SAT solvers and the ability of developers to quickly detect and
identify bugs in their solvers.

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations commonly used in knowledge
compilation, including d-DNNF\@.  We present the CRAT file format that
both captures the structure of a POG and can provide a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof is expressed as a sequence of
clause addition and deletion steps, based on the extended resolution
proof system~\cite{Tseitin:1983}.

We have developed a program that can generate CRAT descriptions from
the d-DNNF graphs generated by the state-of-the-art knowledge compiler
\dfour, as well as a fast CRAT proof checker.  We evaluate the
capabilities of our tool chain using benchmark formulas from the 2022
standard and weighted model competitions.  Our tools are able to
handle all but the largest d-DNNF graphs generated by \dfour.
Moreover, we 
present a checker that has been formally verified
in the \lean{} proof assistant, thereby creating the
first formally verified toolchain for
% weighted and unweighted
model counting.

\section{Related Work}

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{cruz-cade-2017,RAT,wetzler14_drattrim}.  Some progress has also
been made on extending these proof systems to quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2021} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  We compare the performance of
our tool chain to theirs as part of our experimental evluation in
Section~\ref{sect:experimental}.  Our experience with developing a
formally verified proof system has shown us that, even for the well
understood framework of extended resolution, ensuring that the system
is sound can be challenging.

We know of know other work addressing the more general question of
certifying the output of a knowledge compiler.

@@

\begin{definition}[Ring Evaluation Problem (REP)]
  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.

@@

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow 
independently assigning weights $W(x)$ and $W(\obar{x})$ for variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi, w)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum} 
\end{eqnarray}
\end{proposition}

@@

\caption{Example formula (A), its POG representation (B), and its CRAT proof (C), (E), and (F)}

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a terminal node with literal $x$ or $\nodeu_1$ is a
  product node containing $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a terminal node with literal $\obar{x}$ or $\nodeu_0$ is a product node containing $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined, via two proof steps, with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal 
  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CRAT file.
\item If there is a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$, 
\item If there are multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
\end{enumerate}
  We then combine as hints the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A))
to generate the target clause for node $\nodeu$.
\end{enumerate}
Observe that most of these steps involve a polynomial number of
operations per recursive call.  The only step that could have
exponential complexity is when a SAT solver must be invoked to validate a
literal.

@@

Completing the equivalence proof of input formula $\phi$ and its POG
representation with root literal $r$ requires showing that
$\modelset(P) \subseteq \modelset(\phi)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\psi$.  Deleting $C$ requires proving that
$\modelset(\psi) = \modelset(\psi \lor C)$, i.e., that deleting $C$
does not make the formula easier to satisfy.

@@

Each input clause deletion can be done as a single RUP step.  The
proof generator builds a sequence of hints from among the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that does not satisfy $C$ will yield
$\assign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} | \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

Working upward through the graph, node $\nodeu$ is {\em marked} when
the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
program adds $\obar{u}$ to the RUP literals and adds the appropriate
defining clause to the list of hints.  A terminal node with literal
$\lit$ will be marked if $\lit \in C$, with no hint required.  If
product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
added to the hints.  Marking sum node $\nodeu$ requires that its two children are marked.
The first defining
clause for this node (see Table~\ref{tab:defining}(B)) will then be added as a hint.  At the very end, the program
(assuming the reverse implication holds) will attempt to mark root
node $\noder$, which would require $\assign(r) = 0$, yielding a
conflict.

It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
each clause deletion requires a single RUP step having a set of hints that is
bounded by the number of POG nodes.

@@

The performance of forward implication, both in its execution time and
the size of the proof generated by two optimizations described here.
A key feature of these optimizations is that they do not require any changes to the
proof framework---they build on the power of extended resolution to
create new logical structures within the proof.

@@

A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
considered each literal $\lit \in L$ separately,
calling a SAT solver
on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ could not be validated with BCP\@.

Literal grouping allows all literals to be validated with a single call to the SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  This can be done by contradiction using the SAT solver.
In particular, looking at the defining clauses for $\nodev$ (Table
\ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
the first defining clause to simplify to
$C = \obar{\lit}_1 \lor  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
  clauses to become tautologies.  The SAT solver can be invoked on formula
  $\simplify{\psi}{\passign} \cup \{ C \}$ to show that is unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
  input clauses and the defining clauses for $\nodev$.

@@

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.  We can exploit the property
that a shared subgraph in the graph generated by a top-down d-DNNF
compiler arises when the program detects that the simplified set of
clauses arising from one partial assignment to the literals matches that
of a previous partial assignment.  When this d-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and also check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The forward implication proof can exploit this property of shared
subgraphs by constructing and proving a {\em lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma$ must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.  The net effect is to allow $\validate$ to
exploit the graph structure of the POG, invoking $\validate$ on each
node at most once.

The challenge for implementing this strategy is to find a way to
represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CRAT file.  Some may be
unalterered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CRAT file to create
the appropriate clauses.  Consider an {\em argument} clause
$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
define product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
Table~\ref{tab:defining}(A), we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
an input clause, we generate a synthetic clause $C'_j$ with guard literal
$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
while input clauses in $\gamma_{\nodeu}$ are left unchanged.
Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
 will then prove a lemma, given by the target clause
 $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
 showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.
